Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-15 dev/paper/sections/conclusion.qmd

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  conference]{IEEEtran}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{placeins}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Endogenous Macrodynamics in Algorithmic Recourse},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Endogenous Macrodynamics in Algorithmic Recourse}
\author{}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting and focused
on single individuals: given some estimated model the goal is to find
valid counterfactuals for individual instance that fulfill various
desiderata. The ability of such counterfactuals to handle dynamics like
data and model drift remains a largely unexplored research challenge at
this point. There has also been surprisingly little work on the related
question of how the actual implementation of recourse by one individual
may affect other individuals. Through this work we aim to close that gap
by systematizing and extending existing knowledge. We first show that
many of the existing methodologies can be collectively described by a
generalized framework. We then argue that the existing framework fails
to account for a hidden external cost of recourse, that only reveals
itself when studying the endogenous dynamics of recourse at the group
level. Through simulation experiments involving various popular
counterfactual generators and several benchmark datasets, we generate a
total XX million Counterfactual Explanations and study the resulting
domain and model shifts. We find that the induced shifts are substantial
enough to likely impede the applicability Algorithmic Recourse in
practice. Fortunately, we find various potential mitigation strategies
that can be used in combination with existing approaches. Our simulation
framework for studying recourse dynamics is fast and open-sourced.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, sharp corners, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

\hypertarget{sec-intro}{%
\section{Introduction}\label{sec-intro}}

Recent advances in Artificial Intelligence (AI) have propelled its
adoption in scientific domains outside of Computer Science including
Healthcare, Bioinformatics, Genetics and the Social Sciences. While this
has in many cases brought benefits in terms of efficiency,
state-of-the-art models like Deep Neural Networks (DNN) have also given
rise a new type of problem in the context of data-driven
decision-making. They are essentially \textbf{black boxes}: so complex,
opaque and underspecified in the data that it is often impossible to
understand how they actually arrive at their decision without auxiliary
tools. Despite this shortcoming, black-box models have grown in
popularity in recent years and have at times created undesirable
societal outcomes \protect\hyperlink{ref-o2016weapons}{{[}1{]}}. The
scientific community has tackled this issue from two different angles:
while some have appealed for a strict focus on inherently interpretable
models \protect\hyperlink{ref-rudin2019stop}{{[}2{]}}, others have
investigated different ways to explain the behaviour of black-box
models. These two sub-domains can be broadly referred to as
\textbf{interpretable AI} and \textbf{explainable AI} (XAI),
respectively.

Among the approaches to XAI that have recently grown in popularity are
\textbf{Counterfactual Explanations} (CE). They explain how inputs into
a model need to change for it to produce different outputs.
Counterfactual Explanations that involve realistic and actionable
changes can be used for the purpose of \textbf{Algorithmic Recourse}
(AR) to help individuals who face adverse outcomes. An example relevant
to the Social Sciences is consumer credit: in this context, AR can be
used to guide individuals in improving their creditworthiness, should
they have previously been denied access to credit based on an automated
decision-making system. A meaningful recourse recommendation for a
denied applicant could be: \emph{``If your net savings rate had been
10\% of your monthly income instead of the actual 8\%, your application
would have been successful. See if you can temporarily cut down on
consumption.''} In the remainder of this paper we will use both
terminologies - recourse and counterfactual - interchangeably to refer
to situations where counterfactuals are generated with the intent to
provide individual recourse.

Existing work in this field has largely worked in a static setting:
various approaches have been proposed to generate counterfactuals for a
given individual that is subject to some pre-trained model. More recent
work has compared different approaches within this static setting
\protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}}. In this work, we go
one step further and ask ourselves: what happens if recourse is provided
and implemented repeatedly? What types of dynamics are introduced and
how do different counterfactual generators compare in this context?

Research on Algorithmic Recourse has also so far typically addressed the
issue from the perspective of one single individual and has indeed been
referred to as \textbf{individual recourse} in some places. Arguably
though, most real-world applications that warrant Algorithmic Recourse
involve potentially large groups of individuals typically competing for
scarce resources. Our work demonstrates that in such scenarios, choices
made by or for one single individual are likely to affect the broader
collective of individuals in ways that current approaches to AR fail to
account for. More specifically, we argue that a strict focus on
minimizing the private costs faced by individuals may be too narrow an
objective.

Figure~\ref{fig-poc} illustrates this idea for a binary problem
involving a probabilistic classifier and the counterfactual generator
proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}:
the implementation of AR for a subset of individuals leads to a domain
shift (b), which in turn triggers a model shift (c). As this game of
implementing AR and updating the classifier is repeated, the decision
boundary moves away from training samples that were originally in the
target class (d). We refer to these types of dynamics as
\textbf{endogenous} because they are induced by the implementation of
recourse itself. The term \textbf{macrodynamics} is borrowed from the
economics literature and used to describe processes involving whole
groups or societies.

\begin{figure}

{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/poc.png}

}

\caption{\label{fig-poc}Dynamics in Algorithmic Recourse: we have a
simple linear classifier trained for binary classification (a); the
implementation of AR for a random subset of individuals leads to a
noticable domain shift (b); as the classifier is retrained we observe a
corresponding model shift (c); as this process is repeated, the decision
boundary moves away from the target class (d).}

\end{figure}

We think that these types of endogenous dynamics may be problematic and
warrant our attention. Firstly, model shifts may inadvertently change
classification outcomes for individuals who never received and
implemented recourse. Secondly and relatedly, we observe in
Figure~\ref{fig-poc} that as the decision boundary moves in the
direction of the non-target class, counterfactual paths become shorter:
in the consumer credit example, individuals that previously would have
been denied credit based on their input features are suddenly considered
as creditworthy. Average default risk across all borrowers can therefore
be expected to increase. Conversely, lenders that anticipate such
dynamics may choose to refrain from offering recourse (and hence credit)
to more than just a tiny share of individuals. In that latter and
perhaps more likely scenario, the probability of being offered recourse
decreases with every individual that implements recourse: in other
words, the actions of first-movers exert a negative externality on
future would-be borrowers.

To the best of our knowledge this is the first work investigating
endogenous macrodynamics in AR. Our contributions to the state of
knowledge are as follows: firstly, we posit a compelling argument that
calls for a novel perspective on Algorithmic Recourse extending our
focus from single individuals to groups. Secondly, we introduce an
experimental framework extending previous work by
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}},
which enables us to study macrodynamics of Algorithmic Recourse through
simulations that can be fully parallelized. Thirdly, we use this
framework to provide a first in-depth analysis of endogenous recourse
dynamics induced by various popular counterfactual generators including
\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}},
\protect\hyperlink{ref-schut2021generating}{{[}6{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}7{]}},
\protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and
\protect\hyperlink{ref-antoran2020getting}{{[}9{]}}. To this end we
propose a number of novel evaluation metrics that can be used to
quantify and benchmark the macrodynamics introduced by the different
generators. Finally, we also discuss what drives endogenous dynamics and
propose strategies to mitigate them.

The remainder of the paper is structured as follows:
Section~\ref{sec-related} places our work in the broader context of
related literature. Section~\ref{sec-method} posits a generalized
framework for gradient-based Algorithmic Recourse and introduces the
notion of hidden external costs. Section~\ref{sec-method-2} sets out our
experimental framework for modeling endogenous macrodynamics in AR.
Section~\ref{sec-empirical} presents our experimental setup, including
the data, classifiers and counterfactual generators that we have
employed. Section~\ref{sec-empirical-2} presents the results for
synthetic and real-world datasets along with evidence for the
effectiveness of various proposed mitigation strategies. Our findings
are then discussed in the broader context of the literature in
Section~\ref{sec-discussion}, before pointing to some of the limitations
of our work as well as avenues for future research in
Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
concludes.

\hypertarget{sec-related}{%
\section{Background}\label{sec-related}}

In this Section we provide a review of the relevant literature. First,
Section~\ref{sec-related-recourse} discusses the existing research
within the domain of Counterfactual Explanations and Algorithmic
Recourse. Then, Section~\ref{sec-related-shifts} presents some of the
previous work on the measurement of dataset and model shifts.

\hypertarget{sec-related-recourse}{%
\subsection{Algorithmic Recourse}\label{sec-related-recourse}}

A framework for Counterfactual Explanations was first proposed in 2017
by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} and has
served as the baseline for most methodologies that have been proposed
since then. Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some
pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs
\(Y \in \mathcal{Y}\). Then we are interested in minimizing the
cost\footnote{Equivalently, others have referred to this quantity as
  \emph{complexity} or simply \emph{distance}} \(H=h(x\prime)\) incurred
by individual \(x\) when moving to a counterfactual state \(x\prime\)
such that the predicted outcome \(M(x\prime)\) corresponds to some
target outcome \(y^*\):

\begin{equation}\protect\hypertarget{eq-obj}{}{
\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = t
}\label{eq-obj}\end{equation}

For implementation purposes, Equation~\ref{eq-obj} is typically
approximated through regularization:

\begin{equation}\protect\hypertarget{eq-solution}{}{
x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime)
}\label{eq-solution}\end{equation}

In the baseline work
\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, the cost
function \(h: \mathcal{X} \mapsto \mathbb{R}\) is proxied by some
distance metric based on the simple intuition that perturbations of
\(x\) are costly to the individual. For models that are differentiable
and produce smooth predictions, Equation~\ref{eq-solution} can be solved
through gradient descent. This summarizes the approach followed in
\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} which we
shall refer to simply as \textbf{Wachter} - the name of the first author
- in the remainder of this paper.

Many approaches for the generation of Algorithmic Recourse have been
described in the literature since 2017. An October 2020 survey by Karimi
et al.~layed out 60 algorithms that have been proposed since 2014
\protect\hyperlink{ref-karimi2020survey}{{[}10{]}}. Another survey
published around the same time by Verma et al.~described 29 algorithms
\protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}. Different
approaches vary primarily in terms of the objective functions they
impose, how they optimize said objective (from brute force through
gradient-based approaches to graph traversal algorithms), and how the
ensure that certain requirements for CE are met. Regarding the latter,
the literature has produced an extensive list of desiderata each
addressing different needs. To name but a few, we are interested in
generating counterfactuals that close
\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, actionable
\protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}, realistic
\protect\hyperlink{ref-schut2021generating}{{[}6{]}}, sparse, diverse
\protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and if possible
causally founded
\protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.

Efforts so far have largely been directed at improving the quality of
Counterfactual Explanations within a static context: given some
pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\), we are
interested in generating one or multiple meaningful Counterfactual
Explanations for some individual characterized by \(x\). The ability of
Counterfactual Explanations to handle dynamics like data and model
shifts remains a largely unexplored research challenge at this point
\protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}. We have been
able to identify only one recent work that considers the implications of
\textbf{exogenous} domain and model shifts in the context of AR
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. Exogenous shifts
are strictly of external origin. For example, they might stem from data
correction, temporal shifts or geospatial changes
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. The authors of
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} propose ROAR - a
framework for Algorithmic Recourse that evidently improves robustness to
such exogenous shifts.

As mentioned earlier, research has so far also generally focused on
generating counterfactuals for single individuals or instances. We have
been able to identify only one existing work that investigates black-box
model behavior towards a group of individuals
\protect\hyperlink{ref-carrizosa2021generating}{{[}15{]}}. The authors
propose an optimization framework that generates collective
counterfactuals. We provide a motivation for doing so from the
perspective of endogenous macrodynamics of Algorithmic Recourse.

\hypertarget{sec-related-shifts}{%
\subsection{Domain and Model Shifts}\label{sec-related-shifts}}

Much attention has been paid to the detection of dataset shifts --
situations where the distribution of data changes over time. Rabanser et
al.~suggest a framework to detect data drift from a minimal number of
samples through the application of two-sample tests
\protect\hyperlink{ref-rabanser2019failing}{{[}16{]}}. This task is a
generalization of the anomaly detection problem for large datasets,
which aims to answer the question if two sets of samples could have been
generated from the same probability distribution. Numerous approaches to
anomaly detection have been summarized
\protect\hyperlink{ref-chandola2009anomaly}{{[}17{]}}. Another
well-established research topic is that of concept drift -- situations
where external variables influence the patterns between the input and
the output of a model
\protect\hyperlink{ref-widmer1996learning}{{[}18{]}}. For instance, Gama
et al.~offer a review of the adaptive learning techniques which can
handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}19{]}}.
Less previous work is available on the related topic of model drift -
changes in model performance over time. Nelson et al.~review how
resistant different machine learning models are to the model drift
\protect\hyperlink{ref-nelson2015evaluating}{{[}20{]}}. Ackerman et
al.~offer a method to detect changes in model performance when ground
truth is not available
\protect\hyperlink{ref-ackerman2021machine}{{[}21{]}}.

In the context of Algorithmic Recourse, domain and model shifts were
first brought up by the authors behind ROAR
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. In their work
they refer to model shifts as simply any perturbation \(\Delta\) to the
parameters of the model in question: \(M\). While this also sets the
baseline for our analysis here, it is worth noting that in
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} these
perturbations are mechanically introduced. In contrast we are interested
in quantifying model shifts that arise endogenously as part of a dynamic
recourse process. In addition to quantifying the magnitude of shifts
\(\Delta\), we aim to also analyse the characteristics of changes to the
model, such as the position of the decision boundary and the overall
decisiveness of the model. We have not been able to identify previous
work on this topic.

\hypertarget{sec-related-benchmark}{%
\subsection{Benchmarking Counterfactual
Generators}\label{sec-related-benchmark}}

Despite the large and growing number of approaches to counterfactual
search, there have been surprisingly few benchmark studies that compare
different methodologies. This may be partially due to limited software
availability in this space. Recent work has started to address this gap:
firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large
benchmarking study using different algorithmic aproaches and numerous
tabular datasets; secondly,
\protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}} introduce a Python
framework - CARLA - that can be used to apply and benchmark different
methodologies; finally, \texttt{CounterfactualExplanations.jl}
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}}
provides an extensible, fast and language-agnostic implementation in
Julia. Since the experiments presented here involve extensive
simulations, we have relied on and extended the Julia implementation due
to the associated performance benefits. In particular, we have built a
framework on top of \texttt{CounterfactualExplanations.jl} that extends
the functionality from static benchmarks to simulation experiments:
\texttt{AlgorithmicRecourseDynamics.jl}\footnote{The package is
  available from \ldots{}}. The core concepts implemented in that
package reflect what is presented in section Section~\ref{sec-method-2}
of this paper.

\hypertarget{sec-method}{%
\section{Gradient-Based Recourse Revisited}\label{sec-method}}

In this section we first set out a generalized framework for
gradient-based counterfactual search in Section~\ref{sec-method-general}
that encapsulates the various individual recourse methods we have chosen
to use in our experiments. We then introduce the notion of a hidden
external cost in algorithmic recourse and extend the existing framework
to explicitly address this cost in the counterfactual search objective.

\hypertarget{sec-method-general}{%
\subsection{From individual recourse
\ldots{}}\label{sec-method-general}}

We have chosen to focus on gradient-based counterfactual search for two
reasons: firstly, they can be seen as direct descendants of our baseline
method - Wachter; secondly, gradient-based search is particularly
well-suited for differentiable black-box models like deep neural
networks, which we focus on in this work. In particular, we include
include the following generators in our simulation experiments below:
\textbf{REVISE} \protect\hyperlink{ref-joshi2019towards}{{[}7{]}},
\textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}9{]}},
\textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}}
and a greedy approach that relies on probabilistic models
\protect\hyperlink{ref-schut2021generating}{{[}6{]}}. Our motivation for
including these different generators in our analysis, is that they all
offer slightly different approaches to generate meaningful
counterfactuals for differentiable black-box models. We hypothesize that
generating more \textbf{meaningful} counterfactuals should mitigate the
endogenous dynamics illustrated in Figure~\ref{fig-poc} in
Section~\ref{sec-intro}. This intuition stems from the underlying idea
that more meaningful counterfactuals are generated by the same or at
least a very similar data generating process as the training data. All
else equal, counterfactuals that fulfill this basic requirement should
be less prone to trigger domain and model shifts.

As we will see next, all of them can be described by the following
generalized form of Equation~\ref{eq-solution}:

\begin{equation}\protect\hypertarget{eq-general}{}{
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\}
\end{aligned}
}\label{eq-general}\end{equation}

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked
\(K\)-dimensional array of counterfactual states and
\(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
state space to the feature space. In Wachter, the state space is the
feature space: \(f\) is just the identity function and the number of
counterfactuals \(K\) is equal to one. Both REVISE and CLUE search
counterfactuals in some latent embedding \(S \subset \mathcal{S}\)
instead of the feature space directly. The latent embedding is learned
by a separate generative model that is tasked with learning the data
generating process (DGP) of \(X\). In this case \(f\) in
Equation~\ref{eq-general} corresponds to the decoder part of the
generative model, in other words the function that maps back from the
latent embedding to the feature space. Provided the generative model is
well-specified, traversing the latent embedding typically results in
realistic and plausible counterfactuals, because they are implicitly
generated by the (learned) DGP
\protect\hyperlink{ref-joshi2019towards}{{[}7{]}}.

CLUE distinguishes itself from REVISE and other counterfactual
generators in that it aims to minimize the predictive uncertainty of the
model in question, \(M\). To quantify predictive uncertainty the authors
rely on entropy estimates for probabilistic models. The approach
proposed by \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, which
we shall refer to as \textbf{Greedy}, also works with the subclass of
models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce
predictive uncertainty estimates. The authors show that in this setting
the cost function \(h(\cdot)\) in Equation~\ref{eq-general} is redundant
and meaningful counterfactuals can be generated in a fast and efficient
manner through a modified Jacobian-based Saliency Map Attack (JSMA).
Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}6{]}} also
show that by maximizing the predicted probability of \(x^\prime\) being
assigned to target class \(y^*\) we also implicitly minimize predictive
entropy - as in CLUE. In that sense, CLUE can be seen as equivalent to
REVISE in the Bayesian context and we shall therefore refer to both
approaches collectively as \textbf{Latent Space} generators\footnote{In
  fact, there are a number of other recently proposed approaches to
  counterfactual search that also broadly fall in this same category.
  They largely differ with respect to the chosen generative model: for
  example, the generator proposed by
  \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}23{]}} relies
  on normalizing flows.}.

Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}}
distinguishes itself from all other generators considered here in that
it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et
al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} show that
diverse outcomes can in principal be achieved simply rerunning
counterfactual search multiple times using stochastic gradient descent
(or by randomly initializing the counterfactual). In
\protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is
explicitly proxied via Determinantal Point Processes (DDP): the authors
simply introduce DDP as a component of the cost function
\(h(\mathbf{s}^\prime)\) and therebt produce counterfactuals
\(s_1, ... , s_K\) that look as different from each other as possible.
The implementation of DiCE in the our library of choice -
\texttt{CounterfactualExplanations.jl} - uses that exact approach. It is
worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is
constant and therefore does not affect the objective function
Equation~\ref{eq-general}.

\hypertarget{towards-collective-recourse}{%
\subsection{\ldots{} towards collective
recourse}\label{towards-collective-recourse}}

All of the different approaches introduced above tackle the problem of
Algorithmic Recourse from the perspective of one single
individual\footnote{DiCE recognizes that different individuals may have
  different objective functions, but it does not address the
  interdependencies between different individuals.}. To explicitly
address the issue that individual recourse may affect the outcome and
prospect of other individuals, we propose to extend
Equation~\ref{eq-general} as follows:

\begin{equation}\protect\hypertarget{eq-collective}{}{
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
\end{aligned}
}\label{eq-collective}\end{equation}

Here \(h(f(s_k^\prime))\) denotes the proxy for private costs faced by
the individual as before and \(\lambda_1\) governs to what extent that
private cost ought to be penalized. The newly introduced term
\(g(f(s_k^\prime))\) is meant to capture and address external costs
incurred by the collective of individuals in response to changes in
\(\mathbf{s}^\prime\). The underlying concept of private and external
costs is borrowed from Economics and well-established in that field:
when the decisions or actions by some individual market participant
generate external costs, then the market is said to suffer from negative
externalities and considered inefficient
\protect\hyperlink{ref-pindyck2014microeconomics}{{[}24{]}}. We think
that this concept describes the endogenous dynamics of algorithmic
recourse oberserved here very well. As with individual recourse, the
exact choice of \(g(\cdot)\) is not obvious, nor do we intend to provide
a definite answer in this work, if such even exists. That being said, we
do propose a few potential mitigation strategies in
Section~\ref{sec-empirical-2-mitigate}.

\hypertarget{sec-method-2}{%
\section{Modeling Endogenous Macrodynamics in Algorithmic
Recourse}\label{sec-method-2}}

In the following we describe the framework we propose for modeling and
analysing endogenous macrodynamics in Algorithmic Recourse. We first
describe the basic simulations that were generated to produce the
findings in this work and also constitute the core of
\texttt{AlgorithmicRecourseDynamics.jl} - the Julia package we
introduced earlier. The remainder of this section then introduces
various evaluation metrics that can be used to benchmark different
counterfactual generators with respect to how they perform in the
dynamic setting.

\hypertarget{sec-method-2-experiment}{%
\subsection{Simulations}\label{sec-method-2-experiment}}

The dynamics illustrated in Figure~\ref{fig-poc} in
Section~\ref{sec-intro} were generated through a simple experiment that
aims to simulate the process of Algorithmic Recourse in practice. We
begin in the static setting at time time \(t=0\): given some classifier
\(M\) that was pre-trained on data
\(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\) we generate recourse
for a random batch of \(B\) individuals in the non-target class
(\(\mathcal{D}_0\)). Note that we focus our attention on classification
problems, since classification poses the most common practical use-case
for Algorithmic Recourse. In order to simulate the dynamic process, we
suppose that the model \(M\) is retrained following the actual
implementation of recourse in time \(t=0\). Following the update to the
model, we assume that at time \(t=1\) recourse is generated for yet
another random subset of individuals in the non-target class. This
process is repeated for a number of time periods \(T\). To get a clean
read on endogenous dynamics we keep the total population of samples
closed: we allow existing samples to move from factual to counterfactual
states, but do not allow any entirely new samples to enter the
population. The experimental setup is summarized in Algorithm
\ref{algo-experiment}

\begin{algorithm}
\caption{Simulation Experiment}\label{algo-experiment}
\begin{algorithmic}[1]
\Procedure{Experiment}{$M,\mathcal{D},G$}
\State $t\gets 0$
\While{$t<T$}
\State $\mathcal{D}_B \subset \mathcal{D}_0$ \Comment{Sample from $\mathcal{D}_0$.}
\State $\mathcal{D}_B\gets G(\mathcal{D}_B)$ \Comment{Generate counterfactuals.}
\State $M\gets M(\mathcal{D})$ \Comment{Retrain model.}
\EndWhile
\State \textbf{return} $M,\mathcal{D}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

A noteworthy practical consideration is the choice of \(T\) and \(B\).
The higher these values, the more factual instances undergo recourse
throughout the entire experiment. Of course, this is likely to lead to
more pronounced domain and model shifts by time \(T\). At the same time,
it is generally improbable that a very large part of the population
would request an explanation of the algorithm's decisions. In our
experiments, we choose the values such that \(T \cdot B\) corresponds to
the application of recourse on \(\approx50\%\) of the negative instances
from the initial dataset. As we collect data at each time \(t\), we can
also verify the impact of recourse when it is implemented for a smaller
number of individuals.

>>>>

# Concluding Remarks {#sec-conclusion}

This work has revisited and extended some of the most general and defining concepts underlying the literature on Counterfactual Explanations and, in particular, Algorithmic Recourse. We demonstrate that long-held beliefs as to what defines optimality in AR, are too short-sighted to serve as a foundation for applications of recourse in practice. Specifically, we run multiple experiments that simulate the application of recourse in practice using various popular counterfactual generators and find that all of them induce substantial domain and model shifts. We argue that these shifts should be considered as an expected external cost of individual recourse and call for a paradigm shift from individual to collective recourse. By proposing an adapted counterfactual search objective that incorporates this cost, we make that paradigm shift explicit. We show that this modified objective lends itself to mitigation strategies that can be used to effectively decrease the magnitude of induced domain and model shifts. Through our work we hope to inspire future research on this important topic. To this end we have open-sourced all of our code along with a Julia package - `AlgorithmicRecourseDynamics.jl`. The package is built on top of `CounterfactualExplanations.jl` and inherits its extensibility [@altmeyer2022CounterfactualExplanations]. That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.
